---
title: "Lab 6: Advanced* R Programming"
subtitle: "*okay, maybe intermediate"
date: "February 26, 2020"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
# Housekeeping
- Problem set formatting
---
# ggplot of the week
[jakelawlor](https://github.com/jakelawlor/TidyTuesday_JL/blob/master/CodeFiles/Feb.18.20.CO2Food.R) on Github
---
class: center, middle
# R Programming
## And a Disclaimer

---
class: center, middle, inverse
# Defining Functional Programming
---
# Functional Programming
"Programming paradigm" (i.e. object oriented programming)

Languages using Functional Programming are built on functions - tools to reduce duplication

> Simply put, FP is exactly what it sounds like. If you are doing something more than once, it belongs in a function. In FP, functions are the primary method with which you should carry out tasks. All actions are just (often creative) implementations of functions you’ve written.

>-- [Cleaner R Code with Functional Programming](https://towardsdatascience.com/cleaner-r-code-with-functional-programming-adc37931ef7a)

---
# Functional Programming
Programming languages are (broadly) defined by:
- **Behavior** - what the program does
- **Data** - duh

In functional programming...
- **Behavior** handled purely using functions - self-contained pieces of code that accomplish a task
- **Data** is immutable (can't be changed) - functions in FP take in data as input and produce *new* data as ouput
- Functions are treated like data! 

```{r eval = FALSE}
function myFunc(a, b){ # a and b can be functions or data
 #blah
 return x # x can be functions or data
}
```

---
# Characteristics of Functions in FP

- Functions are pure (1. Deterministic - same input = same output, 2. No side effects - cannot touch or change anything outside of it - no global assignment! no writing to disk or displaying to screen!)

- Functions use immutable data

- Functions guarantee referential transparency (you can replace expression with its corresponding value without changing behavior)

- Functions are first-class entities (you can do anything with functions that you can do with vectors - assign them to variables, store in lists, pass as arguments to other functions, create them within functions, and return them as result of a function)

R isn't strictly speaking a functional programming language (think of some counterexamples of the above), but you should still adopt the style in how you write and think about code. 
---
class: center, middle, inverse
# So What? Applications for R Programming

---
# Atomic Vectors - the building block
> First, a note about notation. When you see a reference to a vector, the writers are probably referring to atomic vectors. There is another important data type called a list or generic vector, with (naturally) different semantics. Lists are also vectors, but lists are not atomic vectors.

> Anyway: The atomic vector is the simplest R data type. Atomic vectors are linear vectors of a single primitive type, like an STL Vector in C++. There is no useful literal syntax for this fundamental data type. To create one of these stupid beasts, assign like:

> `a <- c(1,2,3,4)`
> - [aRgh: Atomic vectors](http://arrgh.tim-smith.us/atomic.html) by Tim Smith

---
# Atomic Vectors - the building block
> Haha, what is c()? It is a function, “c” means “concatenate,” and it assembles the vectors you pass into it end-to-end. “But I passed in numerical primitives,” you might think. Wrong! All naked numbers are double-width floating-point atomic vectors of length one. You’re welcome. Consequences of this include:

> - `a`, above, is a double-typed atomic vector.
> - `is.integer(2)` yields FALSE, because 2 is interpreted as a floating-point value. This has implications for testing equality! You can type an integer literal by suffixing L, as in 2400L.
> - `is.integer(as.integer(c(1,2)))` yields TRUE, because you gave it an atomic vector of integer type.

> - [aRgh: Atomic vectors](http://arrgh.tim-smith.us/atomic.html) by Tim Smith
---
# Atomic Vectors - the building block (cont'd)

Other types of atomic vectors:

logical (may contain TRUE, FALSE, NA), integer, double, complex (as in complex numbers; write as 0+0i), character

If you try to place a value of a particular type into an atomic vector of a different type, R will replace either the value you are trying to add or the entire vector to the more permissive type (*without giving you a warning*)

```{r}
a <- c(1L, 2L, 3L); typeof(a)
a[1] = 2; typeof(a)
a[1] = '2'; typeof(a)
```
---
# Circles of Hell in R - Quirks of the Language 
- Falling into floating point trap
- Growing Objects
- Failing to Vectorize
- Over-Vectorizing
- Not Writing Functions
- Doing Global Assignments
- Tripping on Object Orientation
- Believing it does as intended
- Unhelpfully seeking help

[R Inferno](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf)

---
# Growing Objects
> Let’s look at three ways of doing the same task of creating a sequence of
numbers. Method 1 is to grow the object:

> `vec <- numeric(0)`

> `for(i in 1:n) vec <- c(vec, i)`

> Method 2 creates an object of the final length and then changes the values in
the object by subscripting:

> `vec <- numeric(n)`

> `for(i in 1:n) vec[i] <- i`

> Method 3 directly creates the final object:

> `vec <- 1:n`

>  -- [R Inferno](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf)
---
# Growing Objects
Relative time (seconds) to grow a sequence 

| n |grow |subscript |colon operator
|------:|------:|------:|------:|
| 1000 | 0.01 |0.01 |.00006 |
| 10,000 |0.59 |0.09 |.0004 |
| 100,000 | 133.68 | 0.79 | .005 |
| 1,000,000 | 18,718 | 8.10 | .097 |

> You may wonder why growing objects is so slow. It is the computational
equivalent of suburbanization. When a new size is required, there will not be
enough room where the object is; so it needs to move to a more open space.
Then that space will be too small, and it will need to move again. It takes a lot
of time to move house. Just as in physical suburbanization, growing objects can
spoil all of the available space. You end up with lots of small pieces of available
memory, but no large pieces. This is called fragmenting memory. -- [R Inferno](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf)

---
# Vectorization
Not just about avoiding for() loops ;)

Vectorized functions use the complete vector to give you one result based on that whole vector

i.e. sum(),
paste()

```{r}
firstnames <- c("Maddie", "Rae", "Christian")
lastname <- "Pickens"
paste(firstnames,lastname)
```

---
# Vectorization
Vectorized functions are still loops, just written in C!

```{r}
#fast Fourier transform
fft
```


Using vectorisation for performance means finding the existing R function that is implemented in C and most closely applies to your problem.
---
# Functionals
- Also referred to as the `apply()` family; take function as input and return vector as output
- When you run a ply function, everything happens inside that function and nothing changes in your working environment (FP consistent) -a loop would leave the leftover 'i' in your environment
- Note that "tidyverse" functionals are implemented with the [`purrr`](https://purrr.tidyverse.org/) package

---
# Functionals and Functional Programming
Problem: replace 99's in a dataframe with NA
([source](http://adv-r.had.co.nz/Functional-programming.html))
```{r echo = FALSE}
set.seed(1014)
df <- data.frame(replicate(6, sample(c(1:10, -99), 6, rep = TRUE)))
names(df) <- letters[1:6]
df
```

One option (mistake-prone - can you spot them?):
```{r eval = FALSE}
df$a[df$a == -99] <- NA
df$b[df$b == -99] <- NA
df$c[df$c == -98] <- NA
df$d[df$d == -99] <- NA
df$e[df$e == -99] <- NA
df$f[df$g == -99] <- NA
```
---
# Functionals and Functional Programming
Better option: 
```{r}
fix_missing <- function(x) {
  x[x == -99] <- NA
  x
}

df[] <- lapply(df, fix_missing)
```
---
# Functionals and Functional Programming
```{r eval = FALSE}
summary <- function(x) {
 c(mean(x, na.rm = TRUE),
   median(x, na.rm = TRUE),
   sd(x, na.rm = TRUE),
   mad(x, na.rm = TRUE),
   IQR(x, na.rm = TRUE))
}
lapply(df, summary)
```

Versus: (functions in a list!)
```{r eval = FALSE}
summary <- function(x) {
  funs <- c(mean, median, sd, mad, IQR)
  lapply(funs, function(f) f(x, na.rm = TRUE))
}
```
---
# Functions and Functional Programming
Closures: functions written by functions
```{r}
power <- function(exponent) {
  function(x) {
    x ^ exponent
  }
}

square <- power(2)
square(2)
```

---
# When/How to Actually Use Loops
- Don't nest them :)
- Put as much code outside loops as possible, e.g. creating a sequence first and reusing it
- Make number of iterations as small as possible (e.g. iterate over levels of a factor rather than elements)
- Use for: modifying in place (modifying existing dataframe element), recursive relationships (where element depends on previous value), while loops

---
# Global assignments
Don't make them!! Except for a few functions that clearly have side effects, it is expected in R that a function has no side effects - including making global assignments.

---
# Object Orientation
> There are some functions that are generic. Examples include print, plot,
summary. These functions look at the class attribute of their first argument. If
that argument does have a class attribute, then the generic function looks for a
method of the generic function that matches the class of the argument. If such a
match exists, then the method function is used. If there is no matching method
or if the argument does not have a class, then the default method is used.

>Let’s get specific. The lm (linear model) function returns an object of class
"lm". Among the methods for print are print.lm and print.default. The
result of a call to lm is printed with print.lm. The result of 1:10 is printed
with print.default.

> - [R Inferno](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf)
---
# Object Orientation
```{r}
methods(class = 'lm')
```

---
# Resources
- [Advanced R](https://adv-r.hadley.nz/) - a Hadley Wickham book
- [aRgh: A Newcomer's (Angry) Guide to R](http://arrgh.tim-smith.us/) - programmatic pitfalls and quirks of R for people with backgrounds in other languages
- [The R Inferno](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf) - seven deadly sins of R
